Aneuploidy can be an evolutionary diversion\\ on the path to adaptation

### 2024-01-29 ms/ms.tex
format for MBE

\documentclass[12pt]{extarticle}
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}
\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage{longtable}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\usepackage{booktabs}
\usepackage{pgfplotstable} % http://mirrors.ctan.org/graphics/pgf/contrib/pgfplots/doc/pgfplotstable.pdf
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{subcaption}
\captionsetup[subfigure]{labelformat=empty}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{longtable}
\usepackage{gensymb,siunitx}
\usepackage{cleveref}

@@

% NatBib
\usepackage[super,comma]{natbib}
%\usepackage[numbers,super,compress]{natbib}
%\usepackage[round,colon,authoryear]{natbib}

% Title page
% Chromosomal duplication is a transient evolutionary solution to stress
% \title{Chromosomal duplication can be an evolutionary detour on the path to adaptation}
\title{Aneuploidy can be an evolutionary diversion\\ on the path to adaptation}
%\title{Aneuploidy can delay rather than accelerate adaptation}
%\title{Aneuploidy can impede rather than spur adaptation}

@@

%\paragraph{Aneuploidy can lead to adaptation.}
However, aneuploidy can be beneficial under stressful conditions due to the wide range of phenotypes it can produce, some of which are advantageous~\citep{Pavelka2010, Yang2021}.
Indeed, in a survey of 1,011 yeast strains, aneuploidy has been detected in about 19\%\citep{Peter2018}.
Thus, aneuploidy can lead to rapid adaptation in unicellular eukaryotes~\citep{Gerstein2015,Torres2010, Hong2014, Rancati2008}, as well as to rapid growth of somatic tumour cells~\citep{Schvartzman2010, Sheltzer2017}.
For example, aneuploidy in \yeast facilitates adaptation to a variety of stressful conditions like heat and pH~\citep{Yona2012}, copper~\citep{Covo2014, Gerstein2015}, salt~\citep{Dhar2011, Robinson2023}, and nutrient limitation~\citep{Dunham2002, Gresham2008, Avecilla2022}, with similar results in \calbicans~\citep{Yang2021}.
Importantly, aneuploidy can also lead to drug resistance in pathogenic fungi such as \calbicans~\citep{Selmecki2008, Selmecki2010, Gerstein2020} and \cneoformans~\citep{Sionov2010}, which cause candidiasis and meningoencephalitis, respectively.
Although we focus here on aneuploidy, a similar phenomena of adaptation via gene duplication or amplification has been observed in yeast \citep{Lauer2018}, bacteria \citep{Sonti1989}, and DNA viruses \citep{Elde2012}.

@@

%\paragraph{The present study.}
Here, we test the hypothesis that aneuploidy is an evolutionary ``stepping stone'' that facilitates adaptive evolution by genetic mutations~\citep{Yona2015}.
We develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells. 
Our results show that although aneuploidy reached high frequencies in the experimental populations, the majority of cells in the evolved euploid population likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
These suggests that at the lineage level, aneuploidy may be an ``evolutionary diversion'', rather than a ``stepping stone'', on the path to adaptation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Results
\section*{Results}
In the heat-stress experiment of \citet{Yona2012}, four populations of \yeast evolved under \SI{39}{\celsius}.
Aneuploidy reached high frequency (>95\%) in all four experimental repetitions in the first 450 generations.
Two of the repetitions, marked $H2$ and $H4$, carried no large-scale duplications other than a Chromosome III trisomy.
These two repetitions continued to evolve under the same conditions, wherein aneuploidy was eliminated by generation 1,700 and 2,350 in $H4$ and $H2$, respectively.

@@

%%%%%%%% Fig illustration
\begin{figure}[h]
  \centering
    \includegraphics[width=0.7\textwidth]{../figures/model_illustration.pdf}      
  \caption{
    \textbf{Model Illustration.}
    There are four genotypes in our model: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
    Overall there are two possible trajectories from \euwt\; to \eumt.
    Arrows denote transitions between genotypes, with transitions rates $\mu$ for the beneficial mutation rate and $\delta$ for the aneuploidy rate.
    Elevation differences illustrate the expected, rather than the assumed, fitness differences between the genotypes.
   }
  \label{fig:models}
\end{figure}

@@

% without back transitions:
The estimated beneficial mutation rate is $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$ per genome per generation (that is, roughly 3 out of $10^6$ cell divisions produce a mutant cell with a fitness advantage).
From the literature, the mutation rate per base pair is roughly $2-3\cdot10^{-10}$ (refs.~\citep{Zhu2014,Lynch2008}), but it may be higher under heat stress, as several stresses~\citep{Heidenreich2007}, including heat~\citep{Huang2018}, may cause hypermutation in yeast.
If we assume a 10-fold increase over the mutation rate reported in the literature, then the estimated beneficial mutation rate can be explained by a genomic target size of 1,000 base pairs (that is, 1,000 base pairs across the genome in which a mutation would provide a fitness advantage): $3\cdot10^{-10} \times 10 \times 1,000=3\cdot10^{-6}$.
Supporting this, \citet{Jarolim2013} found 279 genes that contributed to survival after a sudden shift from \SI{30}{\celsius} to \SI{50}{\celsius}, and \citet{Flynn2020} used a deep mutational scan of a single protein, Hsp90, to find 465 amino-acid variants (out of 14,160) that significantly increased growth rate in \SI{37}{\celsius}.
Furthermore, \citet{Yona2012} found at least 10 genes on Chromosome III that increased heat tolerance when over-expressed.
Assuming that other chromosomes also have a similar number of heat-tolerance genes (and even more, as Chromosome III is one of the smallest chromosomes~\citep{Gilchrist2019}), we estimate a total of 160 heat-tolerance genes in the genome.
Indeed, mutations were found in 97 genes in an evolutionary experiment with yeast under heat stress~\citep{Huang2018}.
Thus, to get a genomic target size of 1,000, it is enough that the average gene target size (number of base pairs in a gene in which a mutation is beneficial) is 6.25 base pairs. For example, \citet{Kohn2014} found a target size of 11 in a proton exporter gene (\emph{PMA1}) that contributes to high-salt adaptation.

@@

% FIG posterior
%% generated with posterior-plot.ipynb
\begin{figure}[h]
  \centering
\includegraphics[width=0.8\textwidth]{../figures/posterior.pdf}
  \caption{
  \textbf{Posterior distribution of model parameters.}
On the diagonal, the marginal posterior distribution of each model parameter. 
Below the diagonal, the joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively). Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
} 
  \label{fig:posterior}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%

@@

% Fig model comparison 
%% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
\begin{figure}[h!]
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-a.pdf}      
      \label{fig:fit}
  \end{subfigure}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-b.pdf}      
      \label{fig:fit-cumulative}
  \end{subfigure}
  \caption{
    \textbf{Model fit with and without aneuploidy.}
    The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations using MAP parameters of the model with beneficial aneuploidy (blue; $\delta>0, \; w_{\euwt}<w_{\anwt}<w_{\anmt}<w_{\eumt}$) compared to alternative models: a model with the same parameter values but without aneuploidy (gray, $\delta=0$, concentrated at $t=450$); a model fitted to the data assuming no aneuploidy (green, $\delta=0$); a model fitted to the data assuming neutral aneuploidy (yellow, $\delta>0, \; w_{\anwt}=w_{\euwt}, \; w_{\anmt}=w_{\eumt}$); and a model with beneficial aneuploidy and an extended prior distribution (pink).
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450.
Thus, the blue distribution has a better fit compared to the other distributions (the gray distribution has a particulary poor fit).
The MAP likelihood (\cref{eq:heatstress-likelihood}) is 0.84, 0.78, 0.67, and 0.14 for the models represented by blue, yellow, green, and pink distributions, respectively. 
\textbf{(A)} Histogram of the time to fixation of \eumt. The last bin contains all values equal or greater than 3,000.
\textbf{(B)} Cumulative distribution of the time to fixation. 
     }
  \label{fig:2n*-fixation}
\end{figure}

@@

This result is not unique to the MAP parameter estimate. 
We simulated genotype frequency dynamics using parameter samples from the posterior distribution, and computed the posterior distribution of $F_A$ (\Cref{fig:FA}B). 
The posterior mode $F_A$ was just 0.147 [0.0154-0.370 95\% CI] and only in 489 of 100,000 posterior samples (0.489\%) $F_A$ was  larger than 0.5 (see Supporting Material for results when transitions to less-fit genotypes are allowed, such as \eumt\ to \anmt).  
Thus, if we sample a random cell from the evolved \eumt\ population, it is more likely to have descended directly from an euploid cell than from an aneuploid cell. 
The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with both the population size $N$ and the mutation rate, $\mu$ (\Cref{fig:FA}C,D). In some cases it can also be affected by the fitness parameters (\Cref{fig:posterior_FA_supp}).

@@

% Fig dynamics
\begin{figure}[p]
      \centering
      \includegraphics[width=0.8\textwidth]{../figures/FA.pdf}      
  \caption{
    \textbf{Predicted frequency of aneuploid-descended cells.}     
    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: \eumtA\ arose from \anwt, whereas \eumtM\ arose directly from \euwt. Colored curves are 50 simulations using the MAP estimate parameters. Black dashed curves are the expected genotype frequencies without genetic drift (from a deterministic model). 
     See \Cref{fig:MAP_freq_dynamics_log} for log-log scale, in which the sequence of events is easier to observe.
    \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells descended from \anwt\ cells, computed as the average frequency at the end of 100 simulations for 100,000 samples from the parameter posterior distribution.
    Solid and dashed lines show the mode and 95\% CI.
	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axis.
	Yellow star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
	\textbf{(D)} $F_A$ as a function of the population size ($N$, bottom x-axis) and the beneficial mutation rate ($\mu$, top x-axis) in posterior predictions with MAP parameters. 
	Markers show $F_A$ in 250 simulations per population size or mutation rate value.
	Error bars show mean $F_A$ with 95\% CI (bootstrap, $n=10,000$). Blue and red bars for varying population size and mutation rate, respectively.
	Vertical dashed line for population size in the experiment, $6.425\cdot10^6$, and the MAP mutation rate, $2.965\cdot10^{-6}$. Horizontal line for $F_A^{MAP}=0.106$. 
	\textbf{(E)} Population mean fitness in a model without drift using MAP estimate parameters. Solid lines for mean fitness with aneuploidy ($\delta>0$), where the population reaches adaptation (mean fitness at 99.99\% of maximum value) at generation 1,802. Dashed lines for mean fitness without aneuploidy ($\delta=0$), where the population adapts much earlier, at generation 535.
     }
  \label{fig:FA}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\paragraph{Genetic instability in aneuploid cells.}
It has been suggested that aneuploidy increases genomic instability: \citet{Sheltzer2011b} have demonstrated a fold increase of between 2.2 and 7.1 in the mutation rate of disomic yeast (rather than trisomic yeast, the focus of our analysis).
Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 1, 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\Cref{fig:tau}). Furthermore, we computed the WAIC, a criterion for model selection (Methods). The WAIC values were similar for all $\tau$ values (\Cref{table:WAIC}).

@@

Therefore, we do not find evidence of an increase in mutation rate in aneuploid cells. This may be because, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference; or because Chromosome III is one of the smallest chromosomes \citep{Gilchrist2019}.
We also checked the differences in genotype frequency dynamics for different $\tau$ values. We observe $\tau=100$ could be distinguished if accurate data was available for the waiting time until the frequency of \euwt\ to decrease below 95\%~(\Cref{fig:tau-plots}A) or for waiting time for the frequency of \anwt\ to either reach or go below 95\%~(\Cref{fig:tau-plots}B).

Similarly, we did not find evidence for an increase in the rate of chromosome loss in aneuploid cells~\citep{Sheltzer2011b}, probably due to lack of statistical power.
Nevertheless, increasing the rate of chromosome loss (transitions from \anmt\ to \eumt) without increasing the rate of chromosome gain (transitions from \euwt\ to \anwt) increases $F_A$ (\Cref{fig:FA_rates}B), but not to the same extent as increasing the rate of chromosome gain (\Cref{fig:posterior_FA_supp}).
In contrast, increasing the mutation rate in aneuploid cells can have a marked effect on the dynamics: when using the MAP parameter estimates, $F_A$ increases from  0.1 to 0.52 when the mutation rate in aneuploid cells increases 10-fold (\Cref{fig:FA_rates}C).

@@

So how does \eumtM\ prevail?
Initially, the supply rates of \anwt\ and \eumtM\ are $N \delta \approx 11,000$ and $N \mu \approx 19$, respectively (assuming MAP parameter estimates). 
Therefore, both genotypes are expected to appear immediately at the beginning of the experiment (\Cref{fig:MAP_freq_dynamics_log}).
However, \anwt\ appears at a much higher frequency as $\delta \gg \mu$ by 2-3 orders of magnitude.
After they first appear, \eumtM\ has higher fitness.
But as long as the frequency of \euwt\ is high, the supply rate of \anwt\ is higher than that of \eumtM, again due to $\delta \gg \mu$. 
However, supply rates of both genotypes decreases with the frequency of \euwt.
Therefore, when the latter decreases, mainly due to the increase in the frequency of \anwt, both supply rates diminish.
At this stage, the higher fitness of \eumtM\ comes into play and it starts to take over the population, which is mainly composed of \anwt.
For the aneuploid lineage to compete with the mutant lineage, it must produce \eumtA\ via a mutation followed by chromosome loss.
Although this is a stochastic process (due to drift), our results show that the time until \eumtA\ reaches a frequency of ~0.1\% is roughly 450 generations, without much variation (intersection of purple lines and vertical dashed line in \Cref{fig:MAP_freq_dynamics_log}).
However, by that time \eumtM\ is already at a roughly 10-fold higher frequency (1.86\%), and since both mutants have the same fitness, their relative frequency remains roughly the same until the end of the experiment.

@@

% parameter estimates
\paragraph{Rate and fitness effect of aneuploidy and mutation.}
We inferred the rates of aneuploidy and mutation and their effects on fitness. 
We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, higher than a previous estimate of $6.7 \cdot 10^{-6}$ (ref.~~\citep{Zhu2016}). This may be due to genetic instability caused by heat stress~\citep{Chen2012a}, but we note that there is a general scarcity of empirical data on aneuploidy rates.
In addition, we did not find evidence for increased mutation rates in aneuploid cells. 
Previous empirical studies have suggested that genetic instability (e.g., elevated mutation rates) in aneuploid cells is due to stress associated with the aneuploid state~\citep{Bouchonville2009, Chen2012b, Zhu2012, Ippolito2021b}.
However, in the experiment of \citet{Yona2012}, both the wild-type and the aneuploid were under heat stress, which may explain why we didn't find evidence for an increased mutation rate specifically in aneuploid cells.

@@

We use the results of evolutionary experiments reported by \citet{Yona2012}.
In their heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}. 
Aneuploidy fixed (frequency >95\%) in all four population in the first 450 generations. 
Hereafter, fixation or elimination of a genotype by generation $t$ means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier. 
In the original analysis of \citet{Yona2012}, samples were routinely extracted from the evolving populations and tested for indication of heat-shock tolerance. The first generation in which such indication was found was generation 200. Therefore, we determine that aneuploidy did not reach high frequency before generation 200.
The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350 while still under the same conditions of elevated heat (\SI{39}{\celsius}).

>>>>

\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}
\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage{longtable}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\usepackage{booktabs}
\usepackage{pgfplotstable} % http://mirrors.ctan.org/graphics/pgf/contrib/pgfplots/doc/pgfplotstable.pdf
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{subcaption}
\captionsetup[subfigure]{labelformat=empty}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{longtable}
\usepackage{gensymb}
\usepackage{cleveref}

@@

% NatBib
\usepackage[round,colon,authoryear]{natbib}

% Title page
\title{Aneuploidy can be an evolutionary diversion\\ on the path to adaptation}

@@

%\paragraph{Aneuploidy can lead to adaptation.}
However, aneuploidy can be beneficial under stressful conditions due to the wide range of phenotypes it can produce, some of which are advantageous~\citep{Pavelka2010, Yang2021}.
Indeed, in a survey of 1,011 yeast strains, aneuploidy has been detected in about 19\% \citep{Peter2018}.
Thus, aneuploidy can lead to rapid adaptation in unicellular eukaryotes~\citep{Gerstein2015,Torres2010, Hong2014, Rancati2008}, as well as to rapid growth of somatic tumour cells~\citep{Schvartzman2010, Sheltzer2017}.
For example, aneuploidy in \yeast facilitates adaptation to a variety of stressful conditions like heat and pH~\citep{Yona2012}, copper~\citep{Covo2014, Gerstein2015}, salt~\citep{Dhar2011, Robinson2023}, and nutrient limitation~\citep{Dunham2002, Gresham2008, Avecilla2022}, with similar results in \calbicans~\citep{Yang2021}.
Importantly, aneuploidy can also lead to drug resistance in pathogenic fungi such as \calbicans~\citep{Selmecki2008, Selmecki2010, Gerstein2020} and \cneoformans~\citep{Sionov2010}, which cause candidiasis and meningoencephalitis, respectively.
Although we focus here on aneuploidy, a similar phenomena of adaptation via gene duplication or amplification has been observed in yeast \citep{Lauer2018}, bacteria \citep{Sonti1989}, and DNA viruses \citep{Elde2012}.

@@

%\paragraph{The present study.}
Here, we test the hypothesis that aneuploidy is an evolutionary ``stepping stone'' that facilitates adaptive evolution by genetic mutations~\citep{Yona2012}.
We develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells. 
Our results show that although aneuploidy reached high frequencies in the experimental populations, the majority of cells in the evolved euploid population likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
These suggests that at the lineage level, aneuploidy may be an ``evolutionary diversion'', rather than a ``stepping stone'', on the path to adaptation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Results
\section*{Results}
In the heat-stress experiment of \citet{Yona2012}, four populations of \yeast evolved under 39\degree C.
Aneuploidy reached high frequency (>95\%) in all four experimental repetitions in the first 450 generations.
Two of the repetitions, marked $H2$ and $H4$, carried no large-scale duplications other than a Chromosome III trisomy.
These two repetitions continued to evolve under the same conditions, wherein aneuploidy was eliminated by generation 1,700 and 2,350 in $H4$ and $H2$, respectively.

@@

%%%%%%%% Fig illustration
\begin{figure}[h]
  \centering
    \includegraphics[width=0.7\textwidth]{model_illustration.pdf}      
  \caption{
    \textbf{Model Illustration.}
    There are four genotypes in our model: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
    Overall there are two possible trajectories from \euwt\; to \eumt.
    Arrows denote transitions between genotypes, with transitions rates $\mu$ for the beneficial mutation rate and $\delta$ for the aneuploidy rate.
    Elevation differences illustrate the expected, rather than the assumed, fitness differences between the genotypes.
   }
  \label{fig:models}
\end{figure}

@@

% without back transitions:
The estimated beneficial mutation rate is $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$ per genome per generation (that is, roughly 3 out of $10^6$ cell divisions produce a mutant cell with a fitness advantage).
From the literature, the mutation rate per base pair is roughly $2-3\cdot10^{-10}$~\citep{Zhu2014,Lynch2008}, but it may be higher under heat stress, as several stresses~\citep{Heidenreich2007}, including heat \citep{Huang2018}, may cause hypermutation in yeast.
If we assume a 10-fold increase over the mutation rate reported in the literature, then the estimated beneficial mutation rate can be explained by a genomic target size of 1,000 base pairs (that is, 1,000 base pairs across the genome in which a mutation would provide a fitness advantage): $3\cdot10^{-10} \times 10 \times 1,000=3\cdot10^{-6}$.
Supporting this, \citet{Jarolim2013} found 279 genes that contributed to survival after a sudden shift from 30\degree C to 50\degree C, and \citet{Flynn2020} used a deep mutational scan of a single protein, Hsp90, to find 465 amino-acid variants (out of 14,160) that significantly increased growth rate in 37\degree C.
Furthermore, \citet{Yona2012} found at least 10 genes on Chromosome III that increased heat tolerance when over-expressed.
Assuming that other chromosomes also have a similar number of heat-tolerance genes (and even more, as Chromosome III is one of the smallest chromosomes~\citep{Gilchrist2019}), we estimate a total of 160 heat-tolerance genes in the genome.
Indeed, mutations were found in 97 genes in an evolutionary experiment with yeast under heat stress~\citep{Huang2018}.
Thus, to get a genomic target size of 1,000, it is enough that the average gene target size (number of base pairs in a gene in which a mutation is beneficial) is 6.25 base pairs. For example, \citet{Kohn2014} found a target size of 11 in a proton exporter gene (\emph{PMA1}) that contributes to high-salt adaptation.

@@

% FIG posterior
%% generated with posterior-plot.ipynb
\begin{figure}[h]
  \centering
\includegraphics[width=0.8\textwidth]{posterior.pdf}
  \caption{
  \textbf{Posterior distribution of model parameters.}
On the diagonal, the marginal posterior distribution of each model parameter. 
Below the diagonal, the joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively). Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
} 
  \label{fig:posterior}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%

@@

% Fig model comparison 
%% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
\begin{figure}[h!]
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{fixation-plot-a.pdf}      
      \label{fig:fit}
  \end{subfigure}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{fixation-plot-b.pdf}      
      \label{fig:fit-cumulative}
  \end{subfigure}
  \caption{
    \textbf{Model fit with and without aneuploidy.}
    The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations using MAP parameters of the model with beneficial aneuploidy (blue; $\delta>0, \; w_{\euwt}<w_{\anwt}<w_{\anmt}<w_{\eumt}$) compared to alternative models: a model with the same parameter values but without aneuploidy (gray, $\delta=0$, concentrated at $t=450$); a model fitted to the data assuming no aneuploidy (green, $\delta=0$); a model fitted to the data assuming neutral aneuploidy (yellow, $\delta>0, \; w_{\anwt}=w_{\euwt}, \; w_{\anmt}=w_{\eumt}$); and a model with beneficial aneuploidy and an extended prior distribution (pink).
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450.
Thus, the blue distribution has a better fit compared to the other distributions (the gray distribution has a particulary poor fit).
The MAP likelihood (\cref{eq:heatstress-likelihood}) is 0.84, 0.78, 0.67, and 0.14 for the models represented by blue, yellow, green, and pink distributions, respectively. 
\textbf{(A)} Histogram of the time to fixation of \eumt. The last bin contains all values equal or greater than 3,000.
\textbf{(B)} Cumulative distribution of the time to fixation. 
     }
  \label{fig:2n*-fixation}
\end{figure}

@@

This result is not unique to the MAP parameter estimate. 
We simulated genotype frequency dynamics using parameter samples from the posterior distribution, and computed the posterior distribution of $F_A$ (\Cref{fig:FA}B). 
The posterior mode $F_A$ was just 0.147 [0.0154-0.370 95\% CI] and only in 489 of 100,000 posterior samples (0.489\%) $F_A$ was  larger than 0.5 (see Supporting Material for results when transitions to less-fit genotypes are allowed, such as \eumt\ to \anmt).  
Thus, if we sample a random cell from the evolved \eumt\ population, it is more likely to have descended directly from an euploid cell than from an aneuploid cell. 
The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with both the population size $N$ and the mutation rate, $\mu$ (\Cref{fig:FA}C,D). In some cases it can also be affected by the fitness parameters (Figure S10).

@@

% Fig dynamics
\begin{figure}[p]
      \centering
      \includegraphics[width=0.8\textwidth]{FA.pdf}      
  \caption{
    \textbf{Predicted frequency of aneuploid-descended cells.}     
    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: \eumtA\ arose from \anwt, whereas \eumtM\ arose directly from \euwt. Colored curves are 50 simulations using the MAP estimate parameters. Black dashed curves are the expected genotype frequencies without genetic drift (from a deterministic model). 
     See Figure S9 for log-log scale, in which the sequence of events is easier to observe.
    \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells descended from \anwt\ cells, computed as the average frequency at the end of 100 simulations for 100,000 samples from the parameter posterior distribution.
    Solid and dashed lines show the mode and 95\% CI.
	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axis.
	Yellow star shows the MAP estimate. See also Figure S10.
	\textbf{(D)} $F_A$ as a function of the population size ($N$, bottom x-axis) and the beneficial mutation rate ($\mu$, top x-axis) in posterior predictions with MAP parameters. 
	Markers show $F_A$ in 250 simulations per population size or mutation rate value.
	Error bars show mean $F_A$ with 95\% CI (bootstrap, $n=10,000$). Blue and red bars for varying population size and mutation rate, respectively.
	Vertical dashed line for population size in the experiment, $6.425\cdot10^6$, and the MAP mutation rate, $2.965\cdot10^{-6}$. Horizontal line for $F_A^{MAP}=0.106$. 
	\textbf{(E)} Population mean fitness in a model without drift using MAP estimate parameters. Solid lines for mean fitness with aneuploidy ($\delta>0$), where the population reaches adaptation (mean fitness at 99.99\% of maximum value) at generation 1,802. Dashed lines for mean fitness without aneuploidy ($\delta=0$), where the population adapts much earlier, at generation 535.
     }
  \label{fig:FA}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\paragraph{Genetic instability in aneuploid cells.}
It has been suggested that aneuploidy increases genomic instability: \citet{Sheltzer2011b} have demonstrated a fold increase of between 2.2 and 7.1 in the mutation rate of disomic yeast (rather than trisomic yeast, the focus of our analysis).
Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 1, 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (Figure S4). Furthermore, we computed the WAIC, a criterion for model selection (Methods). The WAIC values were similar for all $\tau$ values (Table S1).

@@

Therefore, we do not find evidence of an increase in mutation rate in aneuploid cells. This may be because, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference; or because Chromosome III is one of the smallest chromosomes \citep{Gilchrist2019}.
We also checked the differences in genotype frequency dynamics for different $\tau$ values. We observe $\tau=100$ could be distinguished if accurate data was available for the waiting time until the frequency of \euwt\ to decrease below 95\%~(Figure S5A) or for waiting time for the frequency of \anwt\ to either reach or go below 95\%~(Figure S5B).

Similarly, we did not find evidence for an increase in the rate of chromosome loss in aneuploid cells~\citep{Sheltzer2011b}, probably due to lack of statistical power.
Nevertheless, increasing the rate of chromosome loss (transitions from \anmt\ to \eumt) without increasing the rate of chromosome gain (transitions from \euwt\ to \anwt) increases $F_A$ (Figure S11B), but not to the same extent as increasing the rate of chromosome gain (Figure S10).
In contrast, increasing the mutation rate in aneuploid cells can have a marked effect on the dynamics: when using the MAP parameter estimates, $F_A$ increases from  0.1 to 0.52 when the mutation rate in aneuploid cells increases 10-fold (Figure S11C).

@@

So how does \eumtM\ prevail?
Initially, the supply rates of \anwt\ and \eumtM\ are $N \delta \approx 11,000$ and $N \mu \approx 19$, respectively (assuming MAP parameter estimates). 
Therefore, both genotypes are expected to appear immediately at the beginning of the experiment (Figure S9).
However, \anwt\ appears at a much higher frequency as $\delta \gg \mu$ by 2-3 orders of magnitude.
After they first appear, \eumtM\ has higher fitness.
But as long as the frequency of \euwt\ is high, the supply rate of \anwt\ is higher than that of \eumtM, again due to $\delta \gg \mu$. 
However, supply rates of both genotypes decreases with the frequency of \euwt.
Therefore, when the latter decreases, mainly due to the increase in the frequency of \anwt, both supply rates diminish.
At this stage, the higher fitness of \eumtM\ comes into play and it starts to take over the population, which is mainly composed of \anwt.
For the aneuploid lineage to compete with the mutant lineage, it must produce \eumtA\ via a mutation followed by chromosome loss.
Although this is a stochastic process (due to drift), our results show that the time until \eumtA\ reaches a frequency of ~0.1\% is roughly 450 generations, without much variation (intersection of purple lines and vertical dashed line in Figure S9.
However, by that time \eumtM\ is already at a roughly 10-fold higher frequency (1.86\%), and since both mutants have the same fitness, their relative frequency remains roughly the same until the end of the experiment.

@@

% parameter estimates
\paragraph{Rate and fitness effect of aneuploidy and mutation.}
We inferred the rates of aneuploidy and mutation and their effects on fitness. 
We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, higher than a previous estimate of $6.7 \cdot 10^{-6}$ \citep{Zhu2016}. This may be due to genetic instability caused by heat stress~\citep{Chen2012a}, but we note that there is a general scarcity of empirical data on aneuploidy rates.
In addition, we did not find evidence for increased mutation rates in aneuploid cells. 
Previous empirical studies have suggested that genetic instability (e.g., elevated mutation rates) in aneuploid cells is due to stress associated with the aneuploid state~\citep{Bouchonville2009, Chen2012b, Zhu2012, Ippolito2021b}.
However, in the experiment of \citet{Yona2012}, both the wild-type and the aneuploid were under heat stress, which may explain why we didn't find evidence for an increased mutation rate specifically in aneuploid cells.

@@

We use the results of evolutionary experiments reported by \citet{Yona2012}.
In their heat-stress experiment, four populations of \yeast evolved under 39\degree C. 
Aneuploidy fixed (frequency >95\%) in all four population in the first 450 generations. 
Hereafter, fixation or elimination of a genotype by generation $t$ means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier. 
In the original analysis of \citet{Yona2012}, samples were routinely extracted from the evolving populations and tested for indication of heat-shock tolerance. The first generation in which such indication was found was generation 200. Therefore, we determine that aneuploidy did not reach high frequency before generation 200.
The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350 while still under the same conditions of elevated heat (39\degree C).
